Predicting the effects of character displacement on food-web dynamics

### 2019-03-20 manuscript_ECD_model.Rmd
edited figure legends. added stability boundary text to bifurcation plot

```{r foraging_scenarios, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:foraging_scenarios}**Ecological foraging scenarios.** I examined whether the effect of ecological character displacement on food-web dynamics depended on whether consumers competed for resources that occurred in the same vs. different habitats. In both scenarios, I assumed consumer feeding rate increased linearly with resource density. Later, I relax this assumption and consider a more realistic functional response when resources occurred in different habitats."}
knitr::include_graphics("ECD_model_foraging_scenarios")
```

@@

```{r tradeoff, fig.pos="H", fig.cap="\\label{fig:tradeoff}**Evolutionary tradeoffs in consumer attack rates.** In each foraging scenario, I explored the effects of three different tradeoffs: (i) intermediate combinations of attack rates (a~ii~, a~ij~) are higher than the extremes (green line, n > 1); (ii) extreme combinations of attack rates are higher than intermediate investments (orange line, n < 1); and (iii) all combinations of attack rates have the same total attack rate (blue line, n = 1). Points corresponding to attack rates at the beginning of the simulation. Note that C~i~ was always pre-adapted to R~i~ (a~ii~>a~ij~)."}
knitr::include_graphics("attack_tradeoffs.pdf")
```

@@

```{r Stability, include=FALSE, fig.cap="\\label{fig:Stability}Effect of character displacement on local stability when consumers can (A) and cannot (B) forage for both resources simultaneously. Small and large points correspond to initial and evolved strategies, respectively."}
Stability_Mac <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + #size=5,
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
 #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LS <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r Stability_McCann, include=FALSE, fig.cap="\\label{fig:Stability_McCann}Effect of character displacement on local stability when consumers exhibit a more realistic functional response."}
Stability_McCann <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen)) + 
  geom_line(aes(color=factor(n))) +
  #geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend = FALSE) +
  #geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  geom_point(data = McCann_displacements, aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor, size=Time), color="white", inherit.aes = FALSE) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) +
  #ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_size_manual(values=c(1.5,3)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

# with eco-evo begin and end points
bifurcation_plot <- bind_rows(mutate(sim.df, Competitor = "Yes"), mutate(sim.df_3, Competitor = "No")) %>%
  filter(species == "R") %>%
  ggplot(., aes(x = a11/(a11+a12), y = value, group=interaction(max_min, Competitor))) +
  geom_line(aes(linetype=Competitor), color=cbPalette[2]) +
  geom_point(data = filter(McCann_displacements_bifur, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, shape=Competitor), color="white", fill=cbPalette[2], size=6, inherit.aes = FALSE) +
  geom_point(data = filter(McCann_displacements_bifur, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, shape=Competitor), color="white", fill=cbPalette[2], size=3, inherit.aes = FALSE) +
  scale_shape_manual(values=c(24,21))  +
  scale_linetype_manual(values=c("dotted","solid")) +
  ylab(bquote(atop("Resource abundance","("*R[1]+R[2]*": max and min)"))) +
  xlab(expression(paste("Specialization ", bgroup("(",frac(a[ii],a[ii]+a[ij]),")")))) +
  guides(fill = guide_legend(override.aes=list(shape=21))) +
  geom_segment(x=0.815, xend=0.815, y=0, yend=2.65, arrow = arrow(length = unit(0.03, "npc")), color="grey", fill="grey") +
  theme_cowplot(font_size = 18)
  #geom_vline(xintercept = 0.82, linetype="dotted", color="grey")

@@

Stability_Mac_asymm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements_asymm, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(MacArthur_displacements_asymm, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

Stability_LS_asymm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements_asymm, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(LawlorSmith_displacements_asymm, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

Stability_McCann_asymm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements_asymm, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(McCann_displacements_asymm, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

>>>>

```{r foraging_scenarios, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:foraging_scenarios}**Ecological foraging scenarios.** I examined whether the effect of ecological character displacement on food-web dynamics depended on whether consumers competed for resources that occurred in the same vs. different habitats. In both scenarios, I assumed consumer feeding rate increased linearly with resource density. Later, I relax this assumption and consider a more realistic functional response when resources occurred in different habitats. Note that C~i~ was always pre-adapted to R~i~ (a~ii~ > a~ij~)."}
knitr::include_graphics("ECD_model_foraging_scenarios")
```

@@

```{r tradeoff, fig.pos="H", fig.cap="\\label{fig:tradeoff}**Evolutionary tradeoffs in consumer attack rates.** In each foraging scenario, I explored the effects of three different tradeoffs: intermediate combinations of attack rates (a~ii~, a~ij~) are higher than the extremes (green line, n > 1); extreme combinations of attack rates are higher than intermediate investments (orange line, n < 1); and all combinations of attack rates have the same total attack rate (blue line, n = 1). Points corresponding to attack rates at the beginning of the simulation. Note that C~i~ was always pre-adapted to R~i~ (a~ii~ > a~ij~)."}
knitr::include_graphics("attack_tradeoffs.pdf")
```

@@

```{r Stability, include=FALSE, fig.cap="\\label{fig:Stability}Effect of character displacement on local stability when consumers can (A) and cannot (B) forage for both resources simultaneously. Small and large points correspond to initial and evolved strategies, respectively."}
Stability_Mac <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + #size=5,
  geom_point(data = filter(MacArthur_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
 #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

Stability_LS <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  #xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r Stability_McCann, include=FALSE, fig.cap="\\label{fig:Stability_McCann}Effect of character displacement on local stability when consumers exhibit a more realistic functional response."}
Stability_McCann <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen)) + 
  geom_line(aes(color=factor(n))) +
  #geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend = FALSE) +
  #geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  geom_point(data = McCann_displacements, aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor, size=Time), color="white", inherit.aes = FALSE) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) +
  #ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_size_manual(values=c(1.5,3)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

# with eco-evo begin and end points
bifurcation_plot <- bind_rows(mutate(sim.df, Competitor = "Yes"), mutate(sim.df_3, Competitor = "No")) %>%
  filter(species == "R") %>%
  ggplot(., aes(x = a11/(a11+a12), y = value, group=interaction(max_min, Competitor))) +
  geom_line(aes(linetype=Competitor), color=cbPalette[2]) +
  geom_point(data = McCann_displacements_bifur, aes(x = a11/(a11+a12), y=R1+R2, shape=Competitor, size=Time), color=cbPalette[2], fill=cbPalette[2], inherit.aes = FALSE) +
  #geom_point(data = filter(McCann_displacements_bifur, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, shape=Competitor), color="white", fill=cbPalette[2], size=6, inherit.aes = FALSE) +
  #geom_point(data = filter(McCann_displacements_bifur, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, shape=Competitor), color="white", fill=cbPalette[2], size=3, inherit.aes = FALSE) +
  scale_shape_manual(values=c(24,21))  +
  scale_size_manual(values=c(3,6)) +
  scale_linetype_manual(values=c("dotted","solid")) +
  ylab(bquote(atop("Resource abundance","("*R[1]+R[2]*": max and min)"))) +
  xlab(expression(paste("Specialization ", bgroup("(",frac(a[ii],a[ii]+a[ij]),")")))) +
  #guides(fill = guide_legend(override.aes=list(shape=21))) +
  geom_segment(x=0.815, xend=0.815, y=0, yend=2.65, arrow = arrow(length = unit(0.03, "npc")), color="grey", fill="grey") +
  geom_text(x=0.77, y=2, label="Stability\nboundary", color="grey") +
  theme_cowplot(font_size = 18)
  #geom_vline(xintercept = 0.82, linetype="dotted", color="grey")

@@

Stability_Mac_asymm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), MacArthur_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(MacArthur_displacements_asymm, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(MacArthur_displacements_asymm, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

Stability_LS_asymm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements_asymm, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(LawlorSmith_displacements_asymm, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

Stability_McCann_asymm <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements_asymm, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend=FALSE) + 
  geom_point(data = filter(McCann_displacements_asymm, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))
